---
title: "Statistical Modeling with Time-to-event Mediators and Competing Risks"
output: rmarkdown::html_vignette
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Time-to-event Mediators and Competing Risks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



We follow the framework of Arce Domingo-Relloso *et al.* (2024), who extend the mediational g-formula to accommodate longitudinal mediators and competing events in a single causal framework. In studies where both the mediator and competing risks evolve over time, traditional mediation estimators may fail to correctly identify path-specific effects. The approach implemented in `cmest_pathcomprisk` overcomes this challenge by modeling the structure—exposure, mediator history, competing event history, and survival process—and then evaluating the interventions through the g-formula.

This example illustrates how to use `cmest_pathcomprisk` to estimate path-specific causal effects in settings where:

- the mediator is time-varying,
- the outcome is a time-to-event variable, and
- competing risks may occur prior to the event of interest.

The method follows the framework of Arce Domingo-Relloso et al. (2024), which extends the mediational g-formula to accommodate longitudinal mediators and competing events simultaneously. When both the mediator and the competing event process evolve over time, standard mediation estimators may not correctly separate direct and indirect pathways. The approach implemented in `cmest_pathcomprisk` overcomes this limitation by modeling the full joint longitudinal structure—exposure, mediator history, competing event history, and survival process—and applying the g-formula to evaluate counterfactual interventions.

## Path-specific effects

Under this framework, the total effect of an exposure on a time-to-event outcome is decomposed into four interpretable components.

1. Direct effect (DE)  
   The portion of the exposure effect not operating through either the mediator trajectory or the competing event process. This effect is defined by modifying the exposure while holding both mediator and competing event histories at levels that would occur under a reference exposure value.

2. Indirect effect through the mediator (IEM)  
   The part of the effect transmitted through the entire history of the time-varying mediator, while fixing the competing event process to its reference distribution. This represents how changes in the mediator pathway contribute to the exposure–outcome association.

3. Indirect effect through the competing event (IED)  
   The portion of the effect transmitted through the competing-risk process. Because a competing event can alter the risk set for the primary outcome, this pathway measures how differences in the occurrence or timing of the competing event contribute to the exposure effect.

4. Total effect (TE)  
   The overall effect of the exposure on the outcome when the mediator pathway, the competing event pathway, and any direct mechanisms are all allowed to vary naturally.

Under the identification assumptions presented in Arce Domingo-Relloso et al. (2024), each path-specific effect can be expressed using the longitudinal mediational g-formula. The `cmest_pathcomprisk` function implements these estimands by specifying regression models for the exposure, mediator, competing event process, and time-to-event outcome; generating counterfactual trajectories under fixed exposure values; and computing the corresponding g-formula risk contrasts.

In the remainder of this example, we construct a simplified dataset with a mediator measured at three time points, a competing event indicator also measured at three time points, and a time-to-event outcome. We then demonstrate how to estimate the four path-specific effects using `cmest_pathcomprisk`.


## Estimands

We consider a baseline exposure \(A\), a vector of baseline covariates \(L_0\), a longitudinal mediator
\(\bar M_K = (M_1,\dots,M_K)\), a competing event process \(\bar D_K = (D_1,\dots,D_K)\) and a
time-to-event outcome \(Y\) observed over discrete follow-up intervals \(k=1,\dots,K\). Let
\(\bar S_K = (S_1,\dots,S_K)\) denote the survival indicator up to each interval, where
\(S_k = 1\) if the individual is alive and event-free at time \(k\).

We define counterfactual outcomes under possibly contrary-to-fact interventions on the exposure,
the mediator history, and the survival process, following the path-specific framework of
Domingo-Relloso et al. (2024). For two exposure levels \(a\) (reference) and \(a^\*\) (higher level),
the quantities of interest at time \(k+1\) are effects on the risk difference scale:

- Direct effect (DE): effect of changing \(A\) from \(a\) to \(a^\*\) while holding both the mediator
  and the survival process at the levels they would take under exposure \(a\):
  \[
  DE_{k+1}(a^\*, a)
  = \mathbb{E}\big\{Y_{k+1}(a^\*, \bar M_k(a,\bar S_{k-1}(a)), \bar S_k(a,\bar M_k(a)))\big\}
  - \mathbb{E}\big\{Y_{k+1}(a, \bar M_k(a,\bar S_{k-1}(a)), \bar S_k(a,\bar M_k(a)))\big\}.
  \]

- Indirect effect through the longitudinal mediator (IEM): effect that operates by changing the
  entire mediator trajectory from what it would be under \(A=a\) to what it would be under
  \(A=a^\*\), while forcing the survival process to evolve as under \(A=a\):
  \[
  IEM_{k+1}(a^\*, a)
  = \mathbb{E}\big\{Y_{k+1}(a^\*, \bar M_k(a^\*,\bar S_{k-1}(a)), \bar S_k(a,\bar M_k(a^\*)))\big\}
  - \mathbb{E}\big\{Y_{k+1}(a^\*, \bar M_k(a,\bar S_{k-1}(a)), \bar S_k(a,\bar M_k(a)))\big\}.
  \]

- Indirect effect through the competing event (IED): effect that operates by changing the survival
  and competing risk process from what it would be under \(A=a\) to what it would be under
  \(A=a^\*\), while keeping the mediator trajectory at the level it would take under \(A=a^\*\):
  \[
  IED_{k+1}(a^\*, a)
  = \mathbb{E}\big\{Y_{k+1}(a^\*, \bar M_k(a^\*,\bar S_{k-1}(a^\*)), \bar S_k(a^\*,\bar M_k(a^\*)))\big\}
  - \mathbb{E}\big\{Y_{k+1}(a^\*, \bar M_k(a^\*,\bar S_{k-1}(a)), \bar S_k(a,\bar M_k(a^\*)))\big\}.
  \]

- Total effect (TE): overall effect of changing the exposure from \(a\) to \(a^\*\), allowing both the
  mediator and the competing risk process to evolve under their natural levels for each exposure:
  \[
  TE_{k+1}(a^\*, a)
  = \mathbb{E}\big\{Y_{k+1}(a^\*, \bar M_k(a^\*,\bar S_{k-1}(a^\*)), \bar S_k(a^\*,\bar M_k(a^\*)))\big\}
  - \mathbb{E}\big\{Y_{k+1}(a, \bar M_k(a,\bar S_{k-1}(a)), \bar S_k(a,\bar M_k(a)))\big\}.
  \]

By construction, the three path-specific effects sum to the total effect:
\[
TE_{k+1}(a^\*, a) = DE_{k+1}(a^\*, a) + IEM_{k+1}(a^\*, a) + IED_{k+1}(a^\*, a).
\]

In practice, the function `cmest_pathcomprisk()` reports these effects at a user-specified time
horizon \(t^\*\) (for example, 10 years of follow-up), expressed as differences in cumulative risk
(or in cumulative incidence derived from the additive hazards model). It also returns the
proportion of the total effect explained by the paths through the mediator and competing risk,
\[
Q(t^\*) = \frac{IEM(t^\*) + IED(t^\*)}{TE(t^\*)} \times 100,
\]
interpreted as the percentage of the total effect operating through the mediator trajectory
(including the nested pathway through the competing event).

## Estimation

Under standard identifiability conditions for longitudinal mediation with competing risks
(exchangeability, positivity, consistency and no interference), the counterfactual risks above are
identified by the mediational g-formula. For generic arguments \((a_M, a_S, a_Y)\) indicating the
exposure level carried through the mediator, survival and outcome components, the g-formula
identifies
\[
\phi(a_M, a_S, a_Y)
= \mathbb{E}\{Y_{k+1}(a_Y, \bar M_k(a_M,\bar S_{k-1}(a_S)), \bar S_k(a_S,\bar M_k(a_M)))\}
\]
as a functional of the observed data distribution. The path-specific effects can then be written as
contrasts of \(\phi\), for example
\[
DE_{k+1}(a^\*, a) = \phi(a,a,a^\*) - \phi(a,a,a), \quad
IEM_{k+1}(a^\*, a) = \phi(a^\*,a,a^\*) - \phi(a,a,a^\*),
\]
\[
IED_{k+1}(a^\*, a) = \phi(a^\*,a^\*,a^\*) - \phi(a^\*,a,a^\*), \quad
TE_{k+1}(a^\*, a) = \phi(a^\*,a^\*,a^\*) - \phi(a,a,a).
\]

The function `cmest_pathcomprisk()` implements a parametric mediational g-formula algorithm
tailored to time-varying mediators and competing risks:

1. Choose two exposure levels \(a\) and \(a^\*\) (for example, the 25th and 75th percentiles of the
   exposure distribution) and a time horizon \(t^\*\).

2. Fit sequential regression models for the longitudinal mediator and the competing event at each
   visit \(k = 1,\dots,K\), typically using linear or logistic models of the form
   \[
   M_k \mid \mathcal{H}_{k-1} \sim f_M(M_k \mid M_{k-1}, A, L_0), \quad
   D_k \mid \mathcal{H}_k \sim f_D(D_k \mid M_k, A, L_0),
   \]
   where \(\mathcal{H}_{k-1}\) denotes the observed history up to visit \(k-1\).

3. Transform the data into counting-process format and fit an additive hazards model for the
   time-to-event outcome including the full mediator history, exposure and baseline covariates,
   for example using the `timereg` package:
   \[
   \lambda_Y(t \mid A, \bar M(t), L_0)
   = \beta_0(t) + \beta_A(t) A + \beta_M(t)^\top \bar M(t) + \beta_L(t)^\top L_0.
   \]

4. Using the fitted models, simulate counterfactual trajectories of \(\bar M_K\), \(\bar D_K\) and
   \(\bar S_K\) under each of the required intervention regimes:
   - mediator and survival under exposure \(a\),
   - mediator under \(a^\*\) with survival under \(a\),
   - mediator and survival under \(a^\*\),
   and, if needed, hybrid histories that mix these components. For each regime, predict the
   counterfactual outcome process \(Y(t)\) up to \(t^\*\), either on the rate difference scale (via
   the additive hazards coefficients) or on the survival probability scale (via the model-based
   survival function).

5. Average the simulated outcomes over individuals to obtain empirical estimates of the
   counterfactual risks at \(t^\*\) for each regime. Plug these estimates into the contrasts above
   to obtain \(\widehat{DE}(t^\*)\), \(\widehat{IEM}(t^\*)\), \(\widehat{IED}(t^\*)\) and
   \(\widehat{TE}(t^\*)\), and compute
   \[
   \widehat{Q}(t^\*) = \frac{\widehat{IEM}(t^\*) + \widehat{IED}(t^\*)}{\widehat{TE}(t^\*)} \times 100.
   \]

6. Optionally, repeat the entire procedure over bootstrap resamples to obtain confidence
   intervals for each effect and for the mediated proportion \(Q(t^\*)\).

This algorithm mirrors the multistate mediator framework implemented in
`cmest_multistate`, but extends it to the path-specific decomposition with a longitudinal
mediator nested within the competing risk process, as described in Domingo-Relloso et al. (2024).




## DAG representation

---
```{r}

library(dplyr)
library(ggplot2)
library(CMAverse)
```

```{r}
cmdag(
  outcome  = "Y",
  exposure = "A",
  mediator = c("M1", "M2", "M3"),     # mediator measured at 3 time points
  basec    = "L0",                    # baseline confounder
  postc    = c("D1", "D2", "D3"),     # competing risk indicator at 3 time points
  node     = TRUE,
  text_col = "white"
)
```



```{r}
library(ggdag)
library(dagitty)
library(ggplot2)
library(tibble)

## 1. Manually specify coordinates (ggdag manual style)
coords <- tribble(
  ~name, ~x, ~y,
  "L0",  0,   0,
  "A",   1,   0,
  "M1",  2,   0,
  "M2",  3,   0,
  "M3",  4,   0,
  "Y",   5,   0,
  "D1",  2,  -1,
  "D2",  3,  -1,
  "D3",  4,  -1
)

## 2. Define the DAG structure
dag_longi <- dagify(
  # outcome
  Y  ~ A + L0 + M1 + M2 + M3 + D1 + D2 + D3,

  # mediator process
  M1 ~ A + L0,
  M2 ~ A + L0 + M1,
  M3 ~ A + L0 + M2,

  # competing risk process
  D1 ~ A + L0,
  D2 ~ A + L0 + D1 + M1,
  D3 ~ A + L0 + D2 + M2,

  # exposure model
  A  ~ L0,

  exposure = "A",
  outcome  = "Y",
  coords   = coords
)



## 4. Plot following the ggdag manual style
ggplot(dag_longi, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges() +
  geom_dag_node() +
  geom_dag_text(aes(label = name), vjust = -0.8) +
  theme_dag()


```




## `cmest_pathcomprisk` function

To easily implement this estimation procedure, this feature is available via the `cmest_pathcomprisk()` function. It automates the mediational g-formula for time-varying mediators and competing risks on the hazard scale.

## Simulated data example

We simulate the data using a simulated dataset created by Yuchen Zhang. The exact matching mode utilizes an inner bootstrap loop via the `refit = TRUE` logic to adequately capture variability.

```r
library(CMAverse) # or library(cmcrhazard) / library(cmmed)
library(survival)
library(timereg)
library(data.table)

load("simData.Y10D10.Population.RData")

set.seed(2)  
# Subset the data for demonstration
simData <- simData[sample(1:nrow(simData), size = 10000, replace = FALSE), ]
simData <- as.data.table(simData)
simData[, M_1 := M1]
simData[, M_2 := M2]

# Counting-process helpers the outcome model needs
simData[, `:=`(tstart = 0, tstop = time_to_event,
               time.since.first.exam_1 = 0,
               time.since.first.exam_2 = time_to_event)]

# Unique ID for tmerge / reshape
simData[, id_unique := .I]
simData <- as.data.frame(simData)

# ---- Mediator & Competing risks models (fit on the baseline sample as priors) ----
M1 <- lm(M_1 ~ E + C, data = simData)
M2 <- lm(M_2 ~ E + M_1 + C, data = simData)
mreg_list <- list(M1, M2)

L1 <- glm(D1 ~ E + M_1 + C, data = simData, family = binomial())
L2 <- glm(D2 ~ E + M_2 + C, data = simData, family = binomial())
D_list <- list(L1, L2)

# ---- Transformation to counting-process data for the outcome model ----
df <- survival::tmerge(
  data1 = simData,
  data2 = simData,
  id    = id_unique,
  endpt = event(time_to_event, eventHappened)
)

df_tv <- reshape(
  as.data.frame(simData),
  direction = "long",
  varying   = list(M = c("M_1", "M_2"), time = c("time.since.first.exam_1", "time.since.first.exam_2")),
  v.names = c("M", "time.since.first.exam"),
  times   = c(1, 2),
  idvar   = "id_unique"
)
df_tv <- df_tv[order(df_tv$id_unique, df_tv$time), ]
df <- survival::tmerge(df, df_tv, id = id_unique, M = tdc(time.since.first.exam, M))

Y <- timereg::aalen(
  Surv(tstart, tstop, endpt) ~ const(E) + const(M) + const(C),
  data = df, resample.iid = 1, n.sim = 0
)

# ---- Call the unified analysis function ----
a <- as.numeric(quantile(simData$E, 0.25, na.rm = TRUE))
a_star <- as.numeric(quantile(simData$E, 0.75, na.rm = TRUE))

results <- cmest_pathcomprisk(
  D = D_list, 
  mreg = mreg_list, 
  mvar = c("M_1", "M_2"), 
  yreg = Y, 
  avar = "E", 
  a = a,
  astar = a_star,
  data = simData,
  time_points = c("1", "2"), 
  peryr = 100000, 
  nboot = 1000,
  refit = TRUE,                 # Use Exact Match bootstrapping mode 
  yreg_time = "time_to_event",  
  yreg_event = "eventHappened"
)
```

We can directly inspect the calculated effects:

```r
results
```

```
$DE
[1] "116.04 (50.61, 184.85)"

$IEM
[1] "31.52 (-26.96, 88.81)"

$IED
[1] "-0.36 (-0.67, 0)"

$TE
[1] "147.6 (100.05, 194.07)"

$Q
[1] 21.35 
```

The exposure had a large direct effect on the outcome (DE = 116.0), with more modest and uncertain mediation through the longitudinal mediator (IEM = 31.5) and a negligible competing-risk pathway (IED = -0.36). The total effect (TE = 147.6) reflects that most of the exposure's influence operates directly rather than through mediated biological or competing-risk pathways.
